---
title: "Analyzing smartwatch data metrics"
author: "Darius Alexandru Cocirta"
date: "2022-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The Fitbit dataset consists of data recorded by Fitbit devices – wearable fitness trackers that monitor health and activity. These trackers capture details like steps, distance, calories, heart rate, and sleep patterns. The dataset includes diverse measurements collected from users over time, revealing insights into daily routines, activity levels, sleep quality, and general health.

## About dataset

 The original data set has been taken from https://www.kaggle.com/arashnic/fitbit from Möbius. It contains personal fitness tracker from thirty fitbit users.

## Assessment

- exploring potential correlations between different metrics, such as the relationship between active minutes and heart rate, to uncover possible associations
- Identifying daily and weekly  trends in active minutes, sleep duration, and other key metrics to understand how individuals' behaviors change over timel;
- Investigate sleep-related data to determine average sleep duration, sleep consistency, and any potential connections with other factor

```{r message=FALSE, warning=FALSE, include=FALSE}
library("dplyr")
library("janitor")
library("skimr")
library("here")
library("tidyverse")
library("patchwork")
library("ggthemes")
options(warn = -1)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
mydata = read_csv("dailyActivity_merged.csv")
sleep_data = read_csv("sleepDay_merged.csv")
cal = read_csv("hourlyCalories_merged.csv")
heart_rates = read_csv("heartrate_seconds_merged.csv")
steps = read_csv("hourlySteps_merged.csv")
int = read_csv("hourlyIntensities_merged.csv")
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# activity
mydata$ActivityDate=as.POSIXct(mydata$ActivityDate, format="%m/%d/%Y", tz=Sys.timezone())
# sleep
sleep_data$SleepDay=as.POSIXct(sleep_data$SleepDay, format="%m/%d/%Y", tz=Sys.timezone())

mydata = mydata %>% 
  rename(date = ActivityDate)

sleep_data = sleep_data %>% 
  rename(date = SleepDay)

merged_data = merge(mydata, sleep_data, by = c("Id", "date"))
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# deleting duplicates
merged_data = merged_data %>% 
  distinct() %>% 
  drop_na()
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# changing column names to lowercase
colnames(merged_data)
merged_data <- rename_with(merged_data, tolower)
colnames(merged_data)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# sum total number of activity
merged_data = merged_data %>% 
  mutate(totalactivity = veryactiveminutes + lightlyactiveminutes + fairlyactiveminutes, 
         sleep_hours = round(totalminutesasleep/60, 1))

# categorize users activity per day and their sleep
merged_data <- merged_data %>%
  mutate(user_activity_day = case_when(
    .$totalsteps < 5000 ~ "Sedentary",
    .$totalsteps >= 5000 & .$totalsteps < 7499 ~ "Low active",
    .$totalsteps >= 7500 & .$totalsteps < 9999 ~ "Somewhat active",
    .$totalsteps >= 10000 & .$totalsteps < 12499 ~ "Active",
    .$totalsteps >= 12500 ~ "Very active"
  ), sleepquality = case_when(
    .$totalminutesasleep/60 >= 6 & .$totalminutesasleep/60 <= 8 ~ "Sufficient",
    .$totalminutesasleep/60 < 6 ~ "Insufficient",
    .$totalminutesasleep/60 > 8 ~ "Oversleeping"
  ))
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# adding a new column -> weekday
merged_data <- merged_data %>% 
  mutate(weekday = weekdays(date))

merged_data$weekday <- ordered(merged_data$weekday, 
                                levels = c("Monday", "Tuesday", 
                                           "Wednesday", "Thursday", "Friday", "Saturday", "Sunday"))

# how long it takes for each individual to fall asleep (meaning the difference between total minutes in bed and actual sleep minutes)
merged_data = merged_data %>% 
  mutate(minutes_till_sleep = totaltimeinbed - totalminutesasleep)
```

## Steps

```{r echo=FALSE, warning=FALSE, message = FALSE}
# steps vs very active minutes

s0 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = veryactiveminutes)) + 
  geom_point(mapping = aes(x = totalsteps, y = veryactiveminutes), color = "darkblue") + 
  labs(title = "Steps Vs Very Active Minutes") +xlab("total steps") + ylab("very actvie minutes")

# steps vs fairly active minutes
s1 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = fairlyactiveminutes)) + 
  geom_point(mapping = aes(x = totalsteps, y = fairlyactiveminutes), color = "darkblue") + 
  labs(title = "Steps Vs Fairly Active Minutes") +xlab("total steps") + ylab("Fairly actvie minutes")

# steps vs lightly active minutes
s2 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = lightlyactiveminutes)) + 
  geom_point(mapping = aes(x = totalsteps, y = lightlyactiveminutes), color = "darkblue") + 
  labs(title = "Steps Vs Ligthly Active Minutes") +xlab("total steps") + ylab("ligthly actvie minutes")

# steps vs sedentary minutes
s3 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = sedentaryminutes)) + 
  geom_point(mapping = aes(x = totalsteps, y = sedentaryminutes), color = "darkblue") + 
  labs(title = "Steps Vs Sedentary Minutes") +xlab("total steps") + ylab("sedentary minutes")

# steps vs calories
s4 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = calories)) + 
  geom_point(mapping = aes(x = totalsteps, y = calories), color = "darkblue") + 
  labs(title = "Steps Vs Calories") +xlab("total steps") + ylab("calories")

# steps vs total activity
s5 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = totalsteps, y = totalactivity)) + 
  geom_point(mapping = aes(x = totalsteps, y = totalactivity), color = "darkblue") + 
  labs(title = "Steps Vs Total Activity") +xlab("total steps") + ylab("total activity")

s0 + s1 +s2 + s3 +s4 + s5 
```

## Calories

```{r echo=FALSE, warning=FALSE, message = FALSE}
# very active minutes vs calories
p1 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = veryactiveminutes, y = calories)) + 
  geom_point(mapping = aes(x = veryactiveminutes, y = calories), color = "darkblue") + 
  labs(title = "Very active minutes Vs calories") +xlab("very active minutes") + ylab("calories")

# fairlyactiveminutes vs calories
p2 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = fairlyactiveminutes, y = calories)) + 
  geom_point(mapping = aes(x = fairlyactiveminutes, y = calories), color = "darkblue") + 
  labs(title = "Fairly active minutes Vs calories") +xlab("fairly active minutes") + ylab("calories")

# lightlyactiveminutes vs calories
p3 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = lightlyactiveminutes, y = calories)) + 
  geom_point(mapping = aes(x = lightlyactiveminutes, y = calories), color = "darkblue") + 
  labs(title = "Ligthly active minutes Vs calories") +xlab("Ligthly active minutes") + ylab("calories")

# sedentaryminutes vs calories
p4 = ggplot(data=merged_data) + geom_smooth(mapping = aes(x = sedentaryminutes, y = calories)) + 
  geom_point(mapping = aes(x = sedentaryminutes, y = calories), color = "darkblue") + 
  labs(title = "Sedentary minutes Vs calories") +xlab("sedentary minutes") + ylab("calories")


cal$ActivityHour=as.POSIXct(cal$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
cal = cal %>% 
  separate(ActivityHour, into = c("Date", "Time"), sep = " ")

pl1 = cal %>% 
  group_by(Time) %>% 
  summarise(mean_calories = mean(Calories)) %>% 
  ggplot() + geom_col(mapping = aes(x = Time, y = mean_calories), fill = "darkblue") + 
  theme(axis.text.x=element_text(angle=45))

pl1

(p1 | p2 | p3 | p4)/
  pl1
```

## Sleep

```{r echo=FALSE, warning=FALSE, message = FALSE}
### sleep quality vs user's activity
a1 <- ggplot(data = merged_data) +
  geom_bar(mapping = aes(x = user_activity_day, fill = as.factor(user_activity_day))) +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank()
  ) +
  labs(title = "User Activity Distribution")


a2 = ggplot(data = merged_data) + 
  geom_col(mapping = aes(x = sleepquality, y = user_activity_day, fill = sleepquality)) +
  facet_grid(~user_activity_day) +
  labs(title = "User's activity vs Sleep Quality") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y = element_blank()
  ) +
  scale_fill_viridis_d()

a1/
  a2
```

```{r echo=FALSE, warning=FALSE,message = FALSE}
somn = merged_data %>% 
  group_by(id) %>% 
  summarise(mean_sleep_hours = mean(sleep_hours))

ggplot(data = somn) + geom_col(mapping = aes(x = as.character(id), y = mean_sleep_hours, 
                                             fill = mean_sleep_hours)) + 
  scale_fill_gradient(low = "yellow", high = "red") + 
  geom_hline(yintercept = (mean(somn$mean_sleep_hours))) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

```{r echo=FALSE, warning=FALSE, message = FALSE}
# sleep hours vs total steps
xx = ggplot(data = merged_data) + geom_point(mapping = aes(x = sleep_hours, y = totalsteps), color = "darkblue") + 
  geom_smooth(mapping = aes(x = sleep_hours, y = totalsteps))
# sleep hours vs calories
xy = ggplot(data = merged_data) + geom_point(mapping = aes(x = sleep_hours, y = calories), color = "darkblue") + 
  geom_smooth(mapping = aes(x = sleep_hours, y = calories))
# sleep hours vs total activity
xz = ggplot(data = merged_data) + geom_point(mapping = aes(x = sleep_hours, y = totalactivity), color = "darkblue") + 
  geom_smooth(mapping = aes(x = sleep_hours, y = totalactivity))

xx + xy + xz
```

## Weekly

```{r echo=FALSE, warning=FALSE, message = FALSE}
week_summary <- merged_data %>% 
  group_by(weekday) %>% 
  summarise(mean_sleep = mean(totalminutesasleep), mean_steeps = mean(totalsteps), 
            mean_calories = mean(calories), mean_activity = mean(totalactivity))

b1 = ggplot(data = merged_data) + 
  geom_col(mapping = aes(x = user_activity_day, y = weekday, fill = user_activity_day)) +
  facet_grid(~weekday) +
  labs(title = "User's activity vs Weekday") +
  theme(
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
    axis.text.y = element_blank()
  ) +
  scale_fill_viridis_d()

b2 = merged_data%>%
  ggplot(aes(weekday,fill=user_activity_day)) +
  geom_bar(stat = "count", alpha = 0.6, position="fill", show.legend = T) +
  labs(title= "Fig. 7: User activity by weekday", subtitle= "...by level of activity", x ="Day of week", y= "Percentage of users") +
  scale_fill_viridis_d()


b1/
b2
```
```{r echo=FALSE, warning=FALSE, message = FALSE}
plot_one = ggplot(data = week_summary) + geom_col(mapping = aes(x = weekday, y = mean_sleep), fill = "#006699") + 
  theme(axis.text.x = element_text(angle=45, vjust = 0.5))
plot_two = ggplot(data = week_summary) + geom_col(mapping = aes(x = weekday, y = mean_steeps), fill = "#85e0e0") + 
  theme(axis.text.x = element_text(angle=45, vjust = 0.5))
plot_three = ggplot(data = week_summary) + geom_col(mapping = aes(x = weekday, y = mean_calories), fill = "cyan3") + 
  theme(axis.text.x = element_text(angle=45, vjust = 0.5))

plot_one + plot_two +plot_three 
```
## Daily

```{r echo=FALSE, warning=FALSE, message = FALSE}
steps$ActivityHour=as.POSIXct(steps$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
steps = steps %>% 
  separate(ActivityHour, into = c("Date", "Time"), sep = " ")

int$ActivityHour=as.POSIXct(int$ActivityHour, format="%m/%d/%Y %I:%M:%S %p", tz=Sys.timezone())
int = int %>% 
  separate(ActivityHour, into = c("Date", "Time"), sep = " ")

pl2 = int %>% 
  group_by(Time) %>% 
  summarise(mean_intensity = mean(TotalIntensity)) %>% 
  ggplot() + geom_col(mapping = aes(x = Time, y = mean_intensity), fill = "#FF6666") + 
  theme(axis.text.x=element_text(angle=45))
pl2

pl3 = steps %>% 
  group_by(Time) %>% 
  summarise(mean_steps = mean(StepTotal)) %>% 
  ggplot() + geom_col(mapping = aes(x = Time, y = mean_steps), fill = "#FF6666") + 
  theme(axis.text.x=element_text(angle=45))


(pl1 | pl3) / 
  pl2
```

## Users

```{r echo=FALSE, warning=FALSE, message = FALSE}
#visualizing the distribution of users by hours slept daily 
merged_data %>%
  ggplot(aes(x=sleep_hours, fill=..count..)) +
  geom_histogram(alpha=.8, binwidth=.25) +
  scale_fill_gradient(low="green", high="red") +
  labs(title="Fig. 4: Distribution of users...", subtitle= "...by Total Hours of Sleep", x ="Total Hours of Sleep", y= "Count of users") +
  geom_vline(xintercept=mean(merged_data$sleep_hours), color="black", size=1) +
  annotate("text", x=mean(merged_data$sleep_hours)-0.5, y=10, label="average = 6.96 h", color="black", size=5, angle=90)
summary(merged_data$sleep_hours)
```
```{r echo=FALSE, warning=FALSE, message = FALSE}
#visualizing the distribution of users by steps taken daily
merged_data %>%
  ggplot(aes(x=totalsteps, fill=..count..)) +
  geom_histogram(alpha=.8, binwidth=1000) +
  scale_fill_gradient(low="grey", high="blue") +
  labs(title="Fig. 5: Distribution of users...", subtitle= "...by Total Daily Steps", x ="Total Steps per day", y= "Count of users") +
  geom_vline(xintercept=mean(merged_data$totalsteps), color="black", size=1) +
  annotate("text", x=mean(merged_data$totalsteps)-1400, y=35, label="average = 8515 steps", color="black", size=5, angle=90)
summary(merged_data$totalsteps)
```

```{r echo=FALSE, warning=FALSE, message = FALSE}
#visualizing the distribution of users by calories burned daily
merged_data %>%
  ggplot(aes(x=calories, fill=..count..)) +
  geom_histogram(alpha=.8, binwidth=200) +
  scale_fill_gradient(low="grey", high="darkorange") +
  labs(title="Fig. 6: Distribution of users...", subtitle= "...by Calories burned daily", x ="Total Calories per day", y= "Count of users") +
  geom_vline(xintercept=mean(merged_data$calories), color="black", size=1) +
  annotate("text", x=mean(merged_data$calories)+180, y=110, label="average = 2389 kcal", color="black", size=5, angle=0)
summary(merged_data$calories)
```



















